---
title: "ACLED Conflict Hotspot Atlas, Afghanistan 2017 to 2025"
author: "Sajjad Sharifi"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  warning: false
  message: false
---

## Question

Where in Afghanistan has political violence concentrated between 2017 and 2025, where is it intensifying, and how did the August 2021 regime change reshape the spatial pattern?

## Data

ACLED political violence events, January 2017 to May 2025, retrieved via the official `acledR` client. ACLED's Afghanistan coverage begins January 2017, which constrains the time window. District boundaries from GADM v4.1.

## Setup

```{r setup}
library(dplyr)
library(here)
library(lubridate)
library(geodata)
library(sf)
```

## Load and validate

```{r load-data}
afg <- readRDS(here("data", "raw", "afg_acled_2015_2025.rds"))

cat("Total events:", nrow(afg), "\n")
cat("Date range:",
    as.character(min(as.Date(afg$event_date))),
    "to",
    as.character(max(as.Date(afg$event_date))), "\n")
cat("Distinct provinces:", length(unique(afg$admin1)), "\n")
```

## District boundaries

GADM v4.1 admin-2 polygons give us roughly 400 Afghan districts. The first time this runs it downloads the file (about 5 MB), subsequent runs load from the cached file in `data/raw/`.

```{r load-boundaries}
afg_adm2 <- gadm(
  country = "AFG",
  level   = 2,
  path    = here("data", "raw")
) |>
  st_as_sf()

cat("Districts:", nrow(afg_adm2), "\n")
cat("Columns:", paste(names(afg_adm2), collapse = ", "), "\n")

plot(st_geometry(afg_adm2), border = "grey50", lwd = 0.3,
     main = "Afghanistan, GADM admin-2 districts")
```

## 

## Events by district, pre and post August 2021

Each event has a latitude and longitude. We treat them as points, drop the few with missing coordinates, then do a spatial join so each event is tagged with the district polygon that contains it. We then count events per district in two periods: before the Taliban takeover (15 August 2021) and after.

```{r spatial-join}
library(tidyr)

# Convert events to sf points (drop those without coordinates)
afg_pts <- afg |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  mutate(event_date = as.Date(event_date)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Make sure boundaries are in the same CRS
afg_adm2 <- st_transform(afg_adm2, 4326)

# Spatial join: each event gets the district it falls inside
afg_pts <- st_join(
  afg_pts,
  afg_adm2[, c("GID_2", "NAME_1", "NAME_2")],
  join = st_within
)

# Tag period (Aug 15, 2021 is when Kabul fell)
afg_pts <- afg_pts |>
  mutate(period = if_else(event_date < as.Date("2021-08-15"),
                          "pre_Aug2021", "post_Aug2021"))

# Count events per district per period
district_counts <- afg_pts |>
  st_drop_geometry() |>
  filter(!is.na(GID_2)) |>
  count(GID_2, NAME_1, NAME_2, period, name = "events") |>
  pivot_wider(names_from = period, values_from = events, values_fill = 0)

# Diagnostics
cat("Total events:", nrow(afg_pts), "\n")
cat("Events matched to a district:", sum(!is.na(afg_pts$GID_2)), "\n")
cat("Events with no district match:", sum(is.na(afg_pts$GID_2)), "\n")
cat("Districts with at least one event:", nrow(district_counts), "\n")
```

## 

The map below shows reported events per district in the two periods. Districts in grey-yellow had few events, districts in dark red-black had many. Color uses a square-root scale, which gives small counts visible color while keeping high-count districts (Kabul, Kandahar, Nangarhar) distinguishable. Note the two periods are different lengths (roughly 4.6 years before, 3.8 years after) so raw counts are not directly comparable in magnitude, but the spatial reorganization is the point.

```{r district-map, fig.width=10, fig.height=6}
library(ggplot2)
library(scales)

# Join counts back to district polygons
afg_adm2_counts <- afg_adm2 |>
  left_join(district_counts, by = c("GID_2", "NAME_1", "NAME_2")) |>
  mutate(
    pre_Aug2021  = replace_na(pre_Aug2021, 0),
    post_Aug2021 = replace_na(post_Aug2021, 0)
  )

# Long format for facetting
afg_long <- afg_adm2_counts |>
  pivot_longer(
    cols      = c(pre_Aug2021, post_Aug2021),
    names_to  = "period",
    values_to = "events"
  ) |>
  mutate(period = factor(period,
                         levels = c("pre_Aug2021", "post_Aug2021"),
                         labels = c("Before 15 Aug 2021",
                                    "After 15 Aug 2021")))

# Map
p <- ggplot(afg_long) +
  geom_sf(aes(fill = events), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option    = "rocket",
    direction = -1,
    trans     = "sqrt",
    name      = "ACLED events",
    labels    = label_comma(),
    breaks    = c(0, 100, 500, 1500, 3000, 5000)
  ) +
  facet_wrap(~ period) +
  labs(
    title    = "Reported political violence and protest events, Afghan districts",
    subtitle = "ACLED, 2017 to May 2025, split at the Taliban takeover",
    caption  = "Sources: ACLED, GADM v4.1"
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40")
  )

p

# Save for the portfolio
ggsave(
  here("output", "figures", "events_per_district_prepost_2021.png"),
  p, width = 10, height = 6, dpi = 200
)
```

## 

## Event type composition shifted, not just volume

A simple count says conflict dropped. A composition view says what kind of conflict remained. The chart below compares the share of ACLED event categories within each period. Within-period percentages make the two periods comparable even though their absolute event counts differ by roughly 5x.

```{r event-type, fig.width=9, fig.height=5}
# Event counts and shares by period and type
event_type_breakdown <- afg_pts |>
  st_drop_geometry() |>
  count(period, event_type, name = "events") |>
  group_by(period) |>
  mutate(
    share = events / sum(events) * 100,
    period_total = sum(events)
  ) |>
  ungroup() |>
  mutate(period = factor(period,
                         levels = c("pre_Aug2021", "post_Aug2021"),
                         labels = c("Before 15 Aug 2021",
                                    "After 15 Aug 2021")))

# Print the table for the prose
event_type_breakdown |>
  select(period, event_type, events, share) |>
  arrange(period, desc(share)) |>
  print(n = Inf)

# Bar chart, percentage view
p_types <- ggplot(event_type_breakdown,
                  aes(x = reorder(event_type, share, FUN = max),
                      y = share, fill = period)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("Before 15 Aug 2021" = "#7a1d1d",
                               "After 15 Aug 2021"  = "#1a5490")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = NULL,
    y = "Share of period's events",
    title = "How the conflict composition changed across the regime break",
    subtitle = "Within each period, ACLED event categories as % of total",
    fill = NULL,
    caption = "Pre: ~4.6 years. Post: ~3.8 years. Different denominators, so within-period shares."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.major.y = element_blank()
  )

p_types

ggsave(
  here("output", "figures", "event_type_composition.png"),
  p_types, width = 9, height = 5, dpi = 200
)
```

Three patterns stand out. First, battles fell from roughly 62% of pre-2021 events to 30% post-2021, consistent with the disappearance of the armed insurgency once the Republic collapsed. Second, violence against civilians rose from 4% to 34%, even as absolute volume fell roughly 5x; the regime that emerged is consolidating, not contesting. Third, strategic developments climbed from 3% to 17%, picking up Taliban policy announcements that ACLED logs but were uncommon during active war. Protests also rose from near-zero to about 8%, dominated by women's rights demonstrations after the December 2022 university ban. None of this is causal evidence. It is a description of how event mix shifted at the regime break, and it confirms the structural-break framing: the two periods are different phenomena, not the same conflict at different intensity.

## Limitations

ACLED records reported events. Underreporting is higher in areas with limited media access and during communications disruptions. The August 2021 Taliban takeover changed both the conflict dynamic and the reporting environment, so pre and post 2021 series should not be compared without accounting for that structural break.

## Method

To be filled in: kernel density surfaces, Getis-Ord Gi\* hotspots, ARIMA forecasts.

## Results

To be filled in.
